Title TBD

### 2019-06-29 main/draft.tex
First try

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{Kids First or Marriage First?}
\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
Unplanned pregnancies often trigger marriage for unmarried couples. I document two robust empirical facts. First, women who have their first child before or at the year of marriage are statistically more likely to be divorced in the future than those who marry before they have a child. Second, this difference is more pronounced for women with higher income and educational attainment, although it is less common for them to have their kids before marriage. I estimate a lifecycle model with dynamic bargaining, marriage selection and unplanned pregnancy risks. I show that the first fact can be attributed to household bargaining friction, as women agree to enter lower quality marriages to avoid being a single mother, and  the second fact is consistent with childcare requiring both time and money, as this suggests that the costs of being a single mother are higher for more productive women. Qualitatively, this sheds some light on household formation and dissolution and importance of accounting for heterogeneous time costs. Qualitatively, the model suggests sizable causal effect of unplanned pregnancies on observable divorce rates.
\end{abstract}

\section{Introduction}
As of 2018, [40\%] of new births occurred by unmarried mothers. This number is rising for decades, but, more importantly, it is rather heterogeneous, reaching three-quarters for mothers who are African American, [...] under 25 and [...] for high school or less in education. But even for educated and older women this share is substantial [...]. An increase in non-marital cohabitation is also a documented trend, despite of this, persistent non-marital cohabitation is still not commong: according to National Survey of Family And Growth, the median duration of cohabitation is 22 months, and median duration of cohabitation if it ends by marriage is 21.\footnote{Source: Copen et al, \cite{copen}}. Unmarried unions with children are still pretty rare: according to Payne, 2013\nocite{payne}, only about 3\% of children live in unions with unmarried cohabitating parents (as opposed to 21\% with single mothers).

These numbers have important economic implications: children born by unmarried women, as documented by, for instance Kearney, Levine, 2017, display significantly worse future outcomes in terms of education and income, and this pattern survives controlling for many dimensions. Although it is well understood that children of never married single mothers or children of blended families (when a woman marries someone else than child's father) have generally worse outcomes, even when parents marry each other after they made a baby their marriage is statistically less stable, as I argue in this paper, and the larger chances of marriage dissolution do affect future prospects of the children.

This pairs well with common perception that some couples marry at or shortly after the moment they make a baby. If cohabitation is viewed as uncommitted relationship, unexpected fertility may serve as a trigger to obtain more commitment. Narrowly, we can interpret this as shotgun marriage story, however in the times with large abortion access many unplanned pregnancies are aborted, and cohabitating couples seem to \emph{choose} to keep an unplanned baby. The whole role of marriage changes as it is more socially acceptable to cohabitate, though it is much more conventional to do it before having kids. As highlighted by, for instance, Lafortune, 2018, despite of general decrease of role of marriage in the society, securing investment to children is one of the most important forces that makes people marry each other.

This paper highlights one important aspect of this: I argue that when marriage decision is driven by having an unexpected child, bargaining friction makes marital outcome worse. The source of this friction is the large childcare costs and unequal relative exposure to them: as women bear relatively more childcare costs, they are more likely to enter relatively lower quality marriages as a commitment to share a part of these costs. This can rationalize observable differences in marital outcomes between couples with and without out-of-wedlock births.

I document that females who had their first birth before their first marriage have higher chances of divorce in the future, this is mainly driven by women married to the parents of their children, and this is robust and relatively more pronounced for high education or high income groups. I then provide a model incorporating the bargaining mechanism that can rationalize these differences. I estimate the model parameters to match observable patterns in the data and show what it implies about the role of unequal exposure to childcare costs in observable marriage selection and divorce patterns. Shortly, the model suggests if people could perfectly control timing of their births, general share of divorced people would decrease by [...], especially in high-income groups.

>>>>

\newcommand{\laseq}{\stackrel{\lambda\text{-a.e.}}{=}}
\renewcommand{\d}{\underline}
\renewcommand{\u}{\overline}
\newcommand{\td}{\underline{\theta}}
\newcommand{\tu}{\overline{\theta}}
%\renewcommand{\theenumi}{\alph{enumi}}
%\renewcommand{\labelenumi}{(\theenumi)}
%\renewcommand{\theenumii}{\roman{enumii}}
%\renewcommand{\labelenumii}{\theenumii.}
%\renewcommand{\theenumiii}{\arabic{enumiii}}
%\renewcommand{\labelenumiii}{\theenumiii.}
%\renewcommand{\epsilon}{\varepsilon}
\newcommand{\hneq}{\stackrel{\hn}{=}}
\newcommand{\deq}{\stackrel{d}{=}}
\title{Title TBD}
\author{Egor Kozlov}
\begin{document}
%\begin{center}\textbf{Economics 416-1} \\ \emph{By Egor Kozlov}\end{center}
\maketitle

\begin{abstract}
[WIP] Many couples marry shortly after woman gets pregnant. 
\end{abstract}

\section{Introduction}

[This section is not supposed to be read now :)]

%[Paragraph 1: is shotgun marriage a thing]
Despite growing access to contraception and abortion technologies, situations when couples get married shortly after making a baby are still common. In this paper I study economics of how unexpectedly having a baby affects choices to marry. One can call this economics of shotgun marriage: because of number of reasons, both economic and social/cultural, many couples marry shortly after women gets pregnant, and my conjecture is that pregnancy, often unexpected, triggers their marriage.
%The data suggest that at least 
%This can be widely seen in the data, for instance: [large share of birth of unmarried women, peak at $dT=0$, large share of unplanned pregnancies, etc]
[What data should I show here]

%[Paragraph 1.5: divorce]
The main finding I present here is that, based on American Community Survey data, couples who experience shotgun marriage are characterized with substantially worse marital outcomes, that I refer to as having lower marriage quality. The main outcome I focus is the chance of future divorce, though I also document persistent differences in observable school performance of children and women labor market choices. The differences are economically large and survive controlling for many observable characteristics. After documenting the differences I attempt to study sources of them, and to do this I develop and estimate a structural model that can rationalize selection into lower quality marriages and explain different components of the observable differences, focusing mainly on divorce rates.

%[Paragraph 2: why is it important? Child investment]
Understanding divorce rates of couples with young children is way beyond just being a question of family economics, mainly because of persistent differences in outcomes of divorced parents' children.
%Although is widely accepted that having unmarried parents is correlated with lower early-life and adult-life outcomes, and, more recently, children in blended families (having step-parents) are documented to suffer similar problems; 
Having unmarried parents or growing up in a blended family (with step-parents) are accepted as obvious contributors in worse early and adult life outcomes of children. In this work I show that even children who have both of their parents married may face similar risks if they are born in a shotgun marriage, as marriages are not created equal. These risks happen in particular, but not only because of chances of their parents future divorce. Understanding channels driving divorce rates is important too: in particular, I show that early-life childcare costs are important contributor in generating differences in marriage quality, and the policies focusing on childcare may mitigate selection into lower-quality marriages of those who experienced unplanned pregnancy.

In additional to impact of marital stability on children outcomes, many more dimensions of these issues seem to be important. First, being divorced with a child early in life substantially affects women labor market choices and outcomes. Second, such divorces are presumably [source?] related with lowering re-marriage prospectives and possibly future fertility. Finally, general understanding of interaction between pregnancies and marital selection sheds some lights on internal structure of how households make decisions, provides some insights about how one should treat intra-household structure and bargaining and serves as a validity test to existing household models.

The first thing I do in this paper is documenting persistent differences between two groups of people: those who had their first child before or at the year of their first marriage (kids then marriage, K$\to$M) and those who did this at least one year after (M$\to$K). I show that in both raw comparisons and controlling for many dimensions of composition K$\to$M group has larger share of divorced women. Additionally, although share of K$\to$M is not uniform across population, it is economically large for women with high income or college degrees, and the difference in divorce rates in the largest for them. Loosely, it seems that this timing is relatively more important for wealthier population groups.
%[Paragraph 4: what I do: documenting the main fact]

These difference, however, can hardly have any causal interpretation because of large degree of selection into the groups and many choices people make based on the things we cannot track. This is the main justification to use a structural model to dissemble these choices and draw some sort of causal conclusions. Data provide enough facts to discipline the model, suggesting that childcare costs and their heterogeneity are a major factor in these divorces, if we believe that more productive couples choose or have to spend more resources on raising their children.
%[Paragraph 5: what I do: checks and additional facts]

%[Paragraph 6: what I do: setting up the model]
I use dynamic lifecycle model with income heterogeneity, savings, endogenous marriage, intrahousehold bargaining and divorce, fertility shocks and choices and endogenous margin of childcare costs. Although the structure is complicated, all these elements are important in capturing relevant margins in the data, and combining together the model can replicate divorce, marriage and fertility timing observable in the ACS data pretty realistically. Combining ACS with different supplemental data sources including time use and consumer expenditure surveys allow to obtain stable identification of crucial parameters and reproduce key numbers in the data almost perfectly, and then to use the model to draw conclusions about what contributes to the divorce rates that we observe.

%[Paragraph 7: model: the key decomposition]
The main insight from the model is that causal effect, which definition is non-trivial and discussed later, actually has the magnitude similar to the observable cross-sectional difference. According to the model, the cross-sectional effect can be decomposed onto effects of age composition, marriage length, productivity, assets and lower marriage quality, and the last one stands for the major part of the effect. [This is speculative and to be done]. As additional decomposition, the causal effect itself is majorly driven by lower marriage quality of those couples who marry because they had kids and would not marry otherwise,  who correspond to the concept of compliers in empirical microeconomics literature.

On more conceptual level, the model implies that a major part of dissolving marriages is drive by the threat of being single mother, and this state is particularly bad for more productive women because childcare is time-intensive activity in early child's life. Estimated process for unplanned pregnancies creates sizeable risk to young women, and eliminating chances of unplanned pregnancies dramatically reduces divorce rates in early life. [... ... ... ...]
%[Paragraph 8: model: counterfactuals that implies]

%[Paragraph 9: general conclusions: lower marriage quality]
In general, the main result that I present is the  robust association between unplanned pregnancies of unmarried women and lower quality of their marriage. The economics of this is based on large and heterogeneous costs of having children, that cause women to prefer entering lower quality marriages to avoid bearing childcare costs by themselves. The data supports this pattern: as wealthier households spend both more time and more money on their children, the high complimentarity of these activities suggests that more productive women suffer relaitvely larger losses from being single mothers (or, equivalently, gain more in their marriages relative to their outside options). This paper, therefore, suggests that lack of proper insurance against becoming single mother is the major factor creating bad quality marriages.

%[Paragraph 10: general conclusions: implications for hh bargaining]
From the point of view of the theory of household bargaining, this suggests that non-randomness of marriage selection is important: if people were creating couples randomly, no persistence difference in marriage quality would have been observed. [... is this even necessary]
%it is important to take into account people's outside options when modeling household behavior, and
